Prehensile Manipulation Planning: Modelling, Algorithms and Implementation

### 2021-09-06 hpp.tex
Start applying corrections.

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{{\color{blue} Example of manipulation planning problem.} Top: two UR3 robots with one gripper each (X=red, Y=green, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for instance, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}

Robots in industrial manufacturing are today mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. Automating a task with some variability is however very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

This \paper therefore describes a software platform called {Humanoid Path
  Planner} taylored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modeling of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the above software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found by following this \href{https://humanoid-path-planner.github.io/hpp-doc}{link}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces non-linear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

\subsection{Path planning with non-linear constraints}

Some systems are subject to non-linear constraints. These non-linear
constraints define sub-manifolds of the configuration space the robot
must stay on. For instance, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping a same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another way to sample
configurations on a constrained manifold consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} when this is possible. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on non-linear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference resides in the fact that the nodes built on the
tangent space are not immediately projected on the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} propose a general framework to plan task constrained
motions in the presence of moving obstacle. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with non-linear constraints.

@@

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
non-linear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to this latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We address indeed only prehensile manipulation.

@@

Open-source software platforms are an important tool to make possible fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. The most popular one is undoubtedly OMPL \cite{ompl2012}. OMPL integrates many randomized path planning algorithms and is widely used for teaching. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to non-linear constraints.

@@

One of the main differences between our solution and the previously cited ones resides in the way manipulation constraints are compiled into a graph. To our knowledge none of the previous solutions can handle a variety of problems as large as those exposed in Section~\ref{sec:benchmarks}.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. We expose here only properties that are useful for the following developments.

@@

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the 3 first components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the 4 last components $(x,y,z,w)$ represent unit quaternion $w + x {i} + y {j} + z {k}$.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The 2 first components represent the image of the origin. The 2 last components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ. {\color{blue} Using $(\cos\theta,\sin\theta)$ instead of $\theta$ for $SO(2)$ and $SE(2)$ makes the parameterization continuous when $\theta$ discontinuously switches from $-\pi$ to $\pi$}}
  \label{tab:vector representation}
\end{table}

@@

\section{Non-linear constraints and solvers}\label{sec:constraints}

Some tasks require the robot to enforce some non-linear constraints. Foot contact on the ground for a humanoid robot, center of mass projection on a horizontal plane, gaze constraint are a few examples.

\subsection{Non-linear constraints}

\begin{definition}\textbf{Non-linear constraint.} A non-linear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non zero right hand side for the same function $h$. We define for that parameterized non-linear constraints.
\begin{definition}\textbf{Parameterized non-linear constraint.}
  A parameterized non-linear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

To optimize constraint resolution, we perform variable substitution when possible in order to reduce both the number of variables and the dimension of the resulting implicit constraint. We describe here the method that has been first published in~{\color{blue}Mirabel \textit{et al}}~\cite{MirLam2018}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that stores for each configuration variable, the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraint computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores that elements of $out$ are computed by $E$ and $E$ is inserted in the vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single one. For instance, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

>>>>

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{{\color{blue} Example of manipulation planning problem.} Top: two UR3 robots with one gripper each (X=red, Y=green, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for example, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}

Today, robots in industrial manufacturing are mostly programmed by
hand. They repeat the same motion thousands of times with a high
accuracy. However, automating a task with some variability is very
challenging since it requires more programming effort to integrate
sensors and motion planning in the process. A good example of this difficulty is
the Amazon picking challenge \cite{brock-2018}. The work described in this
paper is a small step to the simplification of industrial process
automation in the presence of some variability, like the variation of
the initial position of some object or the presence of unknown
obstacles. The work only covers the motion planning -- and
more accurately the manipulation planning -- issue. The integration in a whole
process is still under development. We think that it
is important not only to develop algorithms, but also to provide them
within an open-source software platform in order to make the
evaluation and then the integration of those algorithms easier.

Therefore this \paper describes a software platform called {Humanoid Path
  Planner} tailored for manipulation planning in robotics. It can handle many types of robots, from manipulator arms to legged humanoid robots.
The main contributions are:
\begin{itemize}
\item an original and general modelling of prehensile manipulation based on
  nonlinear constraints,
\item an original solver for nonlinear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state-of-the-art development tools and methods.
\item a docker image of the aforementioned software with installation instructions provided with this paper. This image makes the experimental results replicable.
\end{itemize}
{\color{blue} Installation instructions can be found by following this \href{https://humanoid-path-planner.github.io/hpp-doc}{link}.}
This paper extends the work presented in {\color{blue} previous papers }\cite{mirabel-iros2016},\cite{MirLam2018} with the following new material:
\begin{itemize}
\item description of the configuration space as a Cartesian product of Lie groups (Section~\ref{partie:preliminary}),
\item unified and detailed definition of the grasp and placement constraints that are only evoked in {\color{blue}Mirabel \textit{et al}}~\cite{mirabel-iros2016} (Section~\ref{sec:manipulation-problem}),
\item automatic construction of the constraint graph (Section~\ref{sec:manipulation-problem}),
\item the docker image of the software,
\item a description of the software platform (Section~\ref{sec:humanoid-path-planner}),
\item experimental results on several different problems.
\end{itemize}

The \paper is organized as follows. \Partie~\ref{sec:related-work} presents some related work for constrained motion planning and manipulation planning.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces nonlinear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to the software platform implementing the notions introduced in the previous sections. Experimental results are provided for a large variety of problems.

@@

\subsection{Path planning with nonlinear constraints}

Some systems are subject to nonlinear constraints. These nonlinear
constraints define sub-manifolds of the configuration space the robot
must stay on. For example, legged robots that must keep contact with
the ground and enforce quasi-static equilibrium, or multi-arm systems
grasping a same object are subject to this type of constraints. As the
volume of the constrained manifold is usually equal to zero, sampling
random configurations satisfying the constraints is an event of zero
probability. To sample configurations on the constrained manifold,
{\color{blue}Dalibard \textit{et al}}~\cite{DalNakLamLau2009} and {\color{blue}Benrenson
\textit{et al}}~\cite{berenson2011} project random configurations using
a generalization of Newton-Raphson algorithm. Another way to sample
configurations on a constrained manifold consists in expressing some
configuration variables with respect to others~\cite{Cortes2002},
\cite{MirLam2018} whenever this is possible. {\color{blue}Jaillet \textit{et al}}~\cite{atlasRRT} propose
another method based on nonlinear projection. They cover the
constrained manifold by growing an atlas composed of local
charts. This approximation provides a probability distribution that is
closer to the uniform distribution over the manifold than the
projection of a uniform distribution over the configuration
space. {\color{blue}Beobkyoon \textit{et al}}~\cite{tb-rrt2016} propose a variation of the latter paper.  The
main difference resides in the fact that the nodes built on the
tangent space are not immediately projected on the manifold.
{\color{blue}Cefalo \textit{et al}}~\cite{CefOri2019} propose a general framework to plan task constrained
motions in the presence of moving obstacle. {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019}
provide an in-depth review of the various approaches to motion
planning with nonlinear constraints.

@@

Manipulation planning is a particular instance of path planning where
some objects are moved by some robots.
Although several instances of the
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multi-contact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation.
The configuration space of the whole system is constrained by
nonlinear constraints due to the fact that objects cannot move by
themselves and should stay in a stable pose when not grasped by a
robot.  The accessible configuration space is thus a union of
sub-manifolds as defined in the previous section. Each of these
manifolds may moreover be a foliation where each leaf corresponds to a
stable pose of an object or to a grasp of an object by a
gripper. The geometrical structure of the problem has been well
understood for a long time~\cite{AlaSimLau89}. Some specific instances of the
problem have even been addressed recently \cite{vendittelli-2018}.

@@

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210},
\cite{KroBek2015}, \cite{LerPha2015} consists in finding a sequence of
manipulation paths that move some objects from an initial pose to a
final pose. The final configuration of the robot is not specified. A
simplifying assumption is the existence of a monotone solution, that is
a solution where each object is grasped at the most once and is moved from
its initial pose to its final pose \cite{stilman2008planning},
\cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective},
\cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion
planner~\cite{cambon:hal-01976081}, \cite{KaeLoz2013}, \cite{LozKae2014}, \cite{18-toussaint-RSS}.

@@

Our work shares many ideas with {\color{blue}Hauser and Ng-Thow-Hing} \cite{HauNgt2011} where the notion of constraint
graph is present, although not as clearly expressed as in this paper. The main
contribution of our work with respect to the latter paper is that the
constraint graph is built automatically at the cost of a more restricted range
of applications. We address indeed only prehensile manipulation.

@@

Open-source software platforms are an important tool to make possible fair
comparison between algorithms.
Several software platforms are available for motion planning and/or manipulation
planning in the robotics community. The most popular one is undoubtedly OMPL \cite{ompl2012}. OMPL integrates many randomized path planning algorithms and is widely used for teaching. Recently, {\color{blue}Kingston \textit{et al}}~\cite{KinMolKav2019} proposed an extension for systems subject to nonlinear constraints.

@@

One of the main differences between our solution and the previously cited ones resides in the way manipulation constraints are compiled into a graph. To our knowledge none of the previous solutions can handle a variety of problems as large as those described in Section~\ref{sec:benchmarks}.

@@

The joint configuration spaces enumerated in the previous paragraph: $\mathbb{R}^n$, $SO(n)$, and $SE(n)$ are all Lie groups. The group operation is $+$ for $\mathbb{R}^n$, and composition denoted as $"."$ for $SE(n)$. We refer to~{\color{blue}Murray \textit{et al}}~\cite[Appendix A]{LMS94} for a thorough definition of Lie groups. Here we detail only properties that are useful for the following developments.

@@

Therefore elements of $SE(3)$ are represented by a vector in $\reals^7$ where the first 3 components represent the image of the origin (vector $\mathbf{t}$ in Equation~\ref{eq:homogeneous matrix}), the last 4 components $(x,y,z,w)$ represent unit quaternion $w + x {i} + y {j} + z {k}$.

@@

Elements of $SE(2)$ are represented by a vector of dimension 4. The first 2 components represent the image of the origin. The last 2 components represent the cosine and sine of the rotation angle. Therefore the homogeneous matrix associated to $\conf=(q_1,q_2,q_3,q_4)$ is
$$
M = \left(\begin{array}{lll}
  q_3 & -q_4 & q_1 \\
  q_4 &  q_3 & q_2 \\
  0   &  0  & 1
\end{array}\right).
$$
Table~\ref{tab:vector representation} compiles this information.
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
    Lie group type & configuration & velocity\\
    \hline
    $SE(3)$ & $(x_1,x_2,x_3,p_1,\cdots,p_4)$ & $\dconf=(\mathbf{v},\omega)$ \\
            & $\in\reals^7$                 & $\in\reals^6$ \\
    $SE(2)$ & $(x_1,x_2,\cos\theta,\sin\theta)$ & $\dconf=(\mathbf{v},\dot{\theta})$\\
            & $\in\reals^4$                 & $\in\reals^3$ \\
    $SO(3)$ & $(p_1,p_2,p_3,p_4)\in\reals^4$ & $\dconf=\omega\in\reals^3$\\
    $SO(2)$ & $(\cos\theta,\sin\theta)\in\reals^2$ & $\dconf=\dot{\theta}\in\reals$\\
    \hline
  \end{tabular}
  \caption{Main Lie group types and their vector representations. Notice that the dimensions of the configuration representation and of the velocity representation may differ. {\color{blue} Using $(\cos\theta,\sin\theta)$ instead of $\theta$ for $SO(2)$ and $SE(2)$ makes the parameterization continuous when $\theta$ discontinuously switches from $-\pi$ to $\pi$}}
  \label{tab:vector representation}
\end{table}

@@

\section{nonlinear constraints and solvers}\label{sec:constraints}

Some tasks require the robot to enforce some nonlinear constraints. Foot contact on the ground for a humanoid robot, center of mass projection on a horizontal plane, gaze constraint are a few examples.

\subsection{nonlinear constraints}

\begin{definition}\textbf{nonlinear constraint.} A nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and is written
\begin{equation}\label{eq:constraint}
h(\conf) = 0.
\end{equation}
\end{definition}
If the robot is subject to several numerical constraints, $h_1,\cdots,h_k$ with values in $\reals^{m_1}\cdots\reals^{m_k}$, these constraints are equivalent to a single constraint $h$ with values in $\reals^{m}$, where $m=\sum_{i=1}^k m_i$, such that
$$
h(\conf) \triangleq \left(\begin{array}{c} h_1(\conf) \\ \vdots \\ h_k(\conf)
\end{array}\right).
$$
It may be useful to use a non-zero right hand side for the same function $h$. We define for that parameterized nonlinear constraints.
\begin{definition}\textbf{Parameterized nonlinear constraint.}
  A parameterized nonlinear constraint is defined by a {\color{blue} piece-wise} differentiable mapping $h$ from $\CS$ to a vector space $\reals^m$ and by a vector $\mathbf{h}_0$ of $\reals^m$ and is written
  $$
  h(\conf) = \mathbf{h}_0.
  $$
\end{definition}
%% begin impl
{\color{blue} Piece-wise} differentiable mappings are represented by abstract Class\\ \href{https://gepettoweb.laas.fr/hpp/hpp-constraints/doxygen-html/classhpp_1_1constraints_1_1DifferentiableFunction.html}{\texttt{DifferentiableFunction}}.
%% end impl

@@

To optimize constraint resolution, we perform variable substitution whenever possible in order to reduce both the number of variables and the dimension of the resulting implicit constraint. Here we describe the method that has been first published in~{\color{blue}Mirabel \textit{et al}}~\cite{MirLam2018}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that stores for each configuration variable, the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraint computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores that elements of $out$ are computed by $E$ and $E$ is inserted in the vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single one. For example, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

\paragraph{Substitution} When an explicit constraint is not successfully added following Algorithm~\ref{alg:ExplicitConstraintSet}, it is handled as an implicit constraint. Therefore, after inserting implicit and explicit constraints, the solver stores a system of equations equivalent to one explicit and one implicit constraints that we denote by:
\begin{eqnarray}\label{eq:substitution-1}
  h(\conf_{in},\conf_{out}) &=& 0, \\
  \label{eq:substitution-2}
  \conf_{out} &=& f(\conf_{in}), \mbox{ where }\\
  \label{eq:substitution-3}
  {in}\cap{out}=\emptyset.
\end{eqnarray}
Substituting~(\ref{eq:substitution-2}) into~(\ref{eq:substitution-1}), we define an implicit constraint on $\conf_{in}$ only:
$$
\tilde{h}(\conf_{in}) \triangleq h(\conf_{in}, f(\conf_{in})) = 0
$$
The \textit{solver by substitution} applies iteration~(\ref{eq:HierarchicalIterative}) to $\tilde{h}$, instead of $h$.  Therefore we need to compute the Jacobian of  $\tilde{h}$:
$$
\frac{\partial \tilde{h}}{\partial \conf_{in}}  =
\frac{\partial h}{\partial \conf_{in}} + \frac{\partial h}{\partial \conf_{out}}.
\frac{\partial f}{\partial \conf_{in}}.
$$
As the Jacobian of $h$ is provided with the implicit constraint, we need to compute $\frac{\partial f}{\partial \conf_{in}}$. Let us recall that $f$ may be the combination of several compatible explicit constraints. Let us denote by $E$ the mapping from $\CS$ to $\CS$ associated to $f$ by Definition~\ref{def:Explicit}.
Let $J$ denote the $nv\times nv$ Jacobian matrix of $E$. Then $J$ is defined by blocks as follows:
\begin{equation}\label{eq:jacobian-explicit}
\begin{array}{cc}
  J_{{in}\times{in}} = I_{|{in}|} & J_{{in}\times{out}} = 0 \\
  J_{{out}\times{in}} = \frac{\partial f}{\partial \conf_{in}} & J_{{out}\times{out}} = 0
\end{array}
\end{equation}
If $E$ is the composition of several explicit constraints $E_i=(in_i,out_i,f_i)$ of Jacobian $J_i$, $i\in I_{nc}$, for an integer $nc$, then
\begin{equation}\label{eq:jacobian-product}
J = \prod_{i=nc}^{1} J_{i},
\end{equation}
with $J_i$ obtained by expression~(\ref{eq:jacobian-explicit}) after replacing $in$, $out$, and $f$ by $in_i$, $out_i$, and $f_i$.
